---
title: "Enabling access to git from Rstudio"
output: html_document
date: "2026-04-09"
---

# ✅ Step 1: Install Git (only once per computer)
1.1. Windows

1. Download Git from:
https://git-scm.com/download/win
2. Run the installer

Accept the default options


Restart RStudio after installation

1.2. macOS
Option A (recommended):

1. Open Terminal
Type:
`git --version`

If Git is not installed, macOS will prompt you to install Command Line Tools

Option B:

Download installer from https://git-scm.com/download/mac

1.3. Linux
Use your system package manager, for example:
`sudo apt install git`

OR

Follow instructions on https://git-scm.com/install/linux

# ✅ Step 2: Create a GitHub Access Token
1. Go to GitHub
2. Click your profile (top right) → Settings
3. Scroll down → Developer settings
4. Click Personal access tokens → Tokens (classic) (or fine-grained tokens)
5. Click Generate new token

Choose:
- Expiration: e.g., 90 days (recommended)
- Scopes (important):
  - ✅ repo (for private repos)
  - ✅ workflow (optional)

6. Click Generate token
7. ⚠️ Copy the token immediately (you won’t see it again)


# ✅ Step 3: Set Token in RStudio

In RStudio, run:
```{r}
install.packages("usethis")
library(usethis)

usethis::edit_r_environ()
```

Then add this line to the file that opens:

`GITHUB_PAT=your_token_here`

Save and restart RStudio.

# ✅ Step 4: Verify it works

Run:
```{r}
install.packages("gitcreds")
gitcreds::gitcreds_get()
```

Or test with:

```{r}
usethis::gh_token_help()
```

💡 Alternative (recommended, easier)

You can also let R handle it:

```{r}
usethis::create_github_token()
```

This opens GitHub and guides you automatically.
