---
title: "Preprocessing and SummarizedExperiment Generation"
author: "Muskan Kumari(Team1)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Overview

This report describes the preprocessing of AlphaDIA precursor-level output into a protein-level expression matrix and creation of a SummarizedExperiment object.

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(SummarizedExperiment)
```

#Read precursor file

```{r}
precursor <- fread("precursors.tsv")

cat("Original dimensions\n")
dim(precursor)
```

#Remove decoy precursors

```{r}
precursor <- precursor[precursor.decoy == 0]

cat("Dimensions after decoy removal\n")
dim(precursor)

table(precursor$precursor.decoy)
```

#Keep required columns

```{r}
precursor <- precursor[, .(

    precursor.mod_seq_charge_hash,

    raw.name,

    precursor.intensity,

    precursor.qval,

    precursor.score,

    pg.proteins,

    pg.genes

)]

head(precursor)
```

#Quality filtering

```{r}
precursor <- precursor[

    precursor.qval < 0.01 &

    !is.na(precursor.intensity) &

    precursor.intensity > 0

]

cat("Dimensions after filtering\n")
dim(precursor)

summary(precursor$precursor.intensity)
```

#Extract representative protein IDs

```{r}
precursor[, protein_id := sub(";.*","",pg.proteins)]

head(
    precursor[,.(pg.proteins,protein_id)]
)
```

#Read contaminant proteins

```{r}
contaminants <- unique(
    readLines("contaminat_protein_ids.txt")
)

contaminants <- gsub("^REFSEQ:","",contaminants)
contaminants <- gsub("^ENSEMBL:","",contaminants)
contaminants <- gsub("^.*:","",contaminants)

length(contaminants)
```

#Identify contaminants

```{r}
precursor[,contaminant := protein_id %in% contaminants]

table(precursor$contaminant)
```

#Aggregate precursor intensities into proteins

```{r}
protein_matrix <- precursor[
,
.(intensity=sum(precursor.intensity,na.rm=TRUE)),
by=.(protein_id,raw.name)
]

dim(protein_matrix)

head(protein_matrix)
```

#Remove contaminant proteins

```{r}
protein_matrix <- protein_matrix[
!protein_id %in% contaminants
]

dim(protein_matrix)
```

#Save long-format protein matrix

```{r}
fwrite(
protein_matrix,
"protein_matrix_long.tsv",
sep="\t"
)
```

#Convert to wide matrix

```{r}
protein_wide <- dcast(

protein_matrix,

protein_id~raw.name,

value.var="intensity",

fill=0

)

dim(protein_wide)

head(protein_wide[,1:6])
```

#Expression matrix

```{r}
expr <- as.matrix(
protein_wide[,-1]
)

rownames(expr) <- protein_wide$protein_id

dim(expr)
```

#Filter proteins with >50% missing values

```{r}
missing <- rowSums(expr==0)

keep <- missing <= 0.5*ncol(expr)

expr <- expr[keep,]

dim(expr)
```

#Log2 transformation

```{r}
expr <- log2(expr+1)

summary(as.vector(expr))
```

#Read metadata

```{r}
metadata <- fread(
"aplha_dia_metadata.csv"
)

head(metadata)
```

#Check sample names

```{r}
all(colnames(expr) %in% metadata$sample)

all(metadata$sample %in% colnames(expr))
```

#Match metadata

```{r}
metadata <- metadata[
match(
colnames(expr),
metadata$sample
),
]

head(metadata)
```

#Create SummarizedExperiment

```{r}
se <- SummarizedExperiment(

assays=list(intensity=expr),

colData=metadata)

se
```

#Save outputs

```{r}
saveRDS(
se,"protein_SE.rds")

write.csv(assay(se),"protein_matrix.csv")

cat("Files saved successfully.\n")
```

#Session Information

```{r}
sessionInfo()
```

